\title{Data Structures - Recitation 0} 
\author{Combinatorial Review, Induction, and Asymptotic Notations.}
\date{September 2021}
\input{tex/texlib/head}

\section{General Course Information}
\begin{enumerate}
    \item Introduce yourself
    \item course mail: huji.dast.2022a@gmail.com
    \item targilim scoring: 0.85 \( \cdot \) Test + 0.15 \( \cdot \) Average(\(N - 2\)). Typed exercises will get up to 5 bonus points. \ctt{Is it still hold for 2022?.}
\item  introduction to the course: \begin{enumerate}
    \item  It’s important \& fun: \begin{enumerate}
        \item We going to learn some data structures: Heaps, Trees, Hash Tables
        \item And some algorithms: Sorting, Minimal Spanning Tree, Shortest Path
    \end{enumerate}
    \item Doing your homework by yourself is the best way to improve your solving problems skill.
\end{enumerate}
\end{enumerate}

\paragraph{Abstract.} Today we will cover induction, infinite series, and asymptotic notation. These tools will come in handy (in the next couple of weeks) when we want to find the runtime complexity of an algorithm, specifically using the ’Master Theorem’, to give asymptotic bounds for recursion relations, and to prove loop invariants using (finite) induction.

\section{Induction}
\paragraph{What is induction?}
\begin{enumerate}
    \item A mathematical proof technique. It is essentially used to prove that a property \(P(n)\) holds for every natural number \(n\).
    \item The method of induction requires two cases to be proved:
    \begin{enumerate}
        \item The first case, called the base case, proves that the property holds for the first element.
        \item The second case, called the induction step, proves that if the property holds for one natural number, then it holds for the next natural number.
    \end{enumerate}
    \item The domino metaphor. 
\end{enumerate}
\paragraph{The two types of induction, their steps, and why it makes sense} (Strong vs Weak) - Emphasize the change in the induction step.
\paragraph{Example 1. (Weak induction)} Prove that \( \forall n \in  N \)
\( \sum_{i=0}^{n}{i} = \frac{n(n+1)}{2} \).


Proof. Base: For \(n = 1\), \(\sum_{i=0}^{1}{1} = 1 = \frac{(1+1)\cdot 1}{2} \).

Assumption: Assume that the claim holds for \(n\).
Step: 
\begin{equation*}
 \sum_{i=0}^{n+1}{i} = \left( \sum_{i=0}^{n}{i} \right) + n+1 = \frac{n(n+1)}{2} + n + 1 = \frac{n(n+1) + 2\cdot (n+1)}{2} = \frac{(n+1)(n+2)}{2} 
\end{equation*}

\ctt{Finish the induction proof and add alternative proof by counting. I am not sure what is favored, exposing both ways (at the first example) will make clear that induction is only a single proofing tool and surly not the only one. Yet from didactic point of view, it might confuses. }

\paragraph{Example 2. (Strong induction)} Let there be a chocolate bar that consists of \(n\) square chocolate blocks. Then it takes exactly \(n - 1\) snaps to separate it into the \(n\) squares no matter how we split it.


Proof. By strong induction. Base: For \(n = 1\), it is clear that we need \(0\) snaps. Assumption: Assume that for \textbf{every} \(m < n \), this claim holds.


Step: We have in our hand the given chocolate bar with \(n\) square chocolate blocks. Then we may snap it anywhere we like, to get two new chocolate bars: one with some \( k \in [n]\) chocolate blocks and one with \(n - k\) chocolate blocks. From the induction assumption, we know that it takes \(k - 1\) snaps to separate the first bar, and \(n - k - 1\) snaps for the second one. And to sum them up, we got exactly \[ (k - 1) + (n - k - 1) + 1 = n - 1 \] snaps.

\section{SERIES AND CONVERGENCE}

\paragraph{Definition 3.} Let \( \{a_n\}^{\infty}_{n = N} \) be a sequence of real number. The sequence of partial sums corresponding to this sequence is \(S_k = \sum^{k}_{n=N}{a_n} \) an which is also a sequence of real numbers, with indices ranging from \(N\) to \(\infty\).
\paragraph{Definition 4.} Let \( \{a_n\}^{\infty}_{n=N} \) be a sequence of real number. We say that the infinite series \( \sum^{\infty}_{n=N}{a_n} \)  converges when the sequence of partial sums converges, i.e. \(S_{k \rightarrow \infty} \rightarrow L \in \mathbb{R} \). In which case we say \( \sum^{\infty}_{n=N}{a_n} = L \). Otherwise, we say it diverges.

Remark 5. Note that we can always ’pad’ the series with zeros, and this does not alter its convergence status. So we shall talk about series which their initial index is zero (or one).
\paragraph{Example 6.} The Harmonic series \( \sum^{\infty}_{n=1}{\frac{1}{n}}\) diverges.

Proof. Assume by contradiction it doesn’t, then there is a bound \( \lim_{k\rightarrow \infty} S_k = S \in \mathbb{R} \). So \(\lim_{k\rightarrow \infty} S_{2k} = S \in  R \). Thus \( \lim_{k} S_{2k}-S_{k} = 0 \), but \( S_{2k}-S_{k} = \frac{1}{n+1} + ... + \frac{1}{2n} \ge n \cdot \frac{1}{2n} = \frac{1}{2} \)
, which is a contradiction, as we assumed \( \lim_{k\rightarrow \infty} S_{2k}-S_{k} = 0 \).

\paragraph{Proposition 7.} If \( \{ S_k \}^{\infty}_{n=0} \) an converges, then \( a_{n\rightarrow\infty} \rightarrow 0 \). Proof. Consider \( \lim_{n\rightarrow\infty} [S_{n} - S{_n-1}]\).

\paragraph{Example 8.} Let \(a_{n}= q^n\) be a sequence of real numbers, \(q \in \mathbb{R}\).
\begin{enumerate}
    \item If \(q < 1\), then \(S_k\) converges.
    \item If \( q \ge 1\), then \(S_k\) diverges.
\end{enumerate}
Proof.
\begin{enumerate}
    \item Note that \[(1-q)S_{k} = \left(1-q\right)\left(1 + q + q^2 .. q^k \right) = \left(1 + q + q^2 + .. + q^k - q - q^2 - .. - q^{k+1} \right) = 1 - q^{k+1}   \] therefore \(S_k = \frac{1-q^{k+1}}{1-q} \), which    converges for \(k \rightarrow \infty \). That is \( S_k \rightarrow \frac{1}{1-q} \).
    \item Exercise.
\end{enumerate}
\subsection{Convergence Tests}
\paragraph{Theorem 9.} (Direct Comparison Test) Let \( \{a_n\}^{\infty}_{n = N} \),\( \{b_n\}^{\infty}_{n = N} \) be real, non-negative sequences, and assume \( \exsits M \in \mathbb{N}\) s.t. \( \forall n \ge M\), \( a_n \le b_n\). Then:
\begin{enumerate}
    \item If \(\sum^{\infty}_{n=1}{b_n}\) converges, so does \(\sum^{\infty}_{n=1}{a_n}\).
    \item If \(\sum^{\infty}_{n=1}{a_n}\) diverges, so does \(\sum^{\infty}_{n=1}{b_n}\).
\end{enumerate}


Proof. Exercise (Guidance: Show that the partial sums corresponding to the sequence \(b_n\) are larger than the partial sums
corr. to \(a_n\), and use a result from Infi about limits).

\paragraph{Theorem 10.} (Limit comparison Test, generalized) Let \( \{a_n\}^{\infty}_{n = N} \),\( \{b_n\}^{\infty}_{n = N} \) be real, non-negative sequences,
and assume \( \exsits M \in \mathbb{N}, r > 0, R > 0 \)  s.t \( \forall n \ge M, r \le \frac{a_n}{b_n} \le R\). Then either \( \sum_{n=1}^{\infty}{a_n} \)
and \( \sum_{n=1}^{\infty}{b_n} \) both converge or they both diverge.
Proof. Use the previous theorem (direct comparison) with \(rb_n \le a_n \le Rb_n\) along with limit arithmetics.

\paragraph{Example 11.} \( \sum_{n=1}^{\infty}{\frac{1}{n^2}} \) converges.
Proof. Note that \( \frac{1}{n^2} \le \frac{1}{n(n-1)} \), so it is enough to prove that the series 
\( \sum_{n=2}^{\infty}{\frac{1}{n(n-1)}}\) converges by Direct Comparison. Notice that the partial sums are:
\begin{equation*}
    S_k = \sum_{n=2}^{k}{\frac{1}{n(n-1)}} = \sum_{n=2}^{k}{\frac{1}{n-1} - \frac{1}{n}} = \\ \left(1 - \frac{1}{2} \right) + \left( \frac{1}{2} - \frac{1}{3} \right) + ... + \left( \frac{1}{k-1} - \frac{1}{k} \right) = \\ 1 - \frac{1}{k}   
\end{equation*}
so the sequence of partial sums clearly converges (to 1).
Proof. (Alternative) Look at the sides of the squares. We could always fit one more line of 2
k
squares for elements 2
k until 2
k+1 − 1 of the
series. The total area of the rect angle here is 2.

% \begin{abstract}
%     Combinatorial Review, Induction, and Asymptotic Notations. 
% \end{abstract}

\section{Asymptotic Notations}
\paragraph{Definition 12.} Let \( f, g : \mathbb{N} \rightarrow \mathbb{R}^{+} \). We say that \( f(n) = O(g(n))\)  if \( \exists N \in \mathbb{N}, \exists c > 0 \) s.t. \( \forall n \ge N \ : \ f(n) \le c \cdot g(n)\)
\paragraph{Example 13.} For exmaple, if \(f(n) = n + 10 \) and \( g(n) = n^2\)
, then \(f(n) = O(g(n)) \) (Draw the graphs) for \(n \ge 5 \):
\(f(n) = n + 10 \le n + 2n = 3n \le n \cdot n = n^2\)
\paragraph{Definition 14.} Let \( f, g : \mathbb{N} \rightarrow \mathbb{R} \)
We say that \(f(n) = \Omega(g(n))\) if \(g(n) = O(f(n))\), equivalently,
\( \exists N \in \mathbb{N}, \exists c > 0 \) s.t. \( \forall n \ge N c_0· g(n) \le f(n) \)
\paragraph{Example 15.} Also if \( f(n) = 5n\) and \(g(n) = n^2\), then \(f(n) = O(g(n))\) (Now discuss intuition - no matter how much we “stretch” \(f, g\) is still the winner)
\paragraph{Definition 16.} Let \( f, g : \mathbb{N} \rightarrow \mathbb{R} \), We say that \(f(n) = \Omega(g(n))\) if:
\( \exists N \in \mathbb{N}, \exists c > 0 \) s.t \( \forall n \ge N \ f(n) \ge c \cdot g(n) \).

\paragraph{Example 17.} For exmaple, if \(f(n) = n + 10\) and \(g(n) = n^2\)
, then \(g(n) = \Omega(f(n))\)

\paragraph{Definition 18.} Let \( f, g : \mathbb{N} \rightarrow \mathbb{R} \), We say that \(f(n) = \Theta(g(n))\) if:
\(f(n) = O(g(n))\) and \(f(n) = \Omega(g(n))\)
That is, we say that \(f(n) = \Theta(g(n))\) if:
\( \exists N \in \mathbb{N}, \exists c_1, c_2 > 0\) s.t. \(\forall n \ge N \ c_1\cdotg(n) \le f(n) \le c_2 \cdot g(n)\)

\paragraph{Example 19.} For every \(f : \mathbb{N} \rightarrow \mathbb{R}, f(n) = \Theta(f(n))\)
\paragraph{Example 20.} If \(p(n) = n^5\) and \(q(n) = 0.5n^5 + n\), then \(p(n) = \Theta(q(n))\)
But why is this example true? This next Lemma helps for intuition:
\paragraph{Lemma 21.} \( \lim_{n \rightarrow \infty} \frac{f(n)}{g(n)} < \infty \Rightarrow f(n) = O(g(n)) \)
Proof. Assume that \(l = \lim_{n \rightarrow \infty} \frac{f(n)}{g(n)} < \infty\). Then for some \(N \in \mathbb{N}\) we have that for all \(n \ge N\):
\( \frac{f(n)}{g(n)} < l + 1 \Rightarrow f(n) < (l + 1) · g(n) \)
Which is exactly what we wanted.
1
\(c_0\)
in this definition is \(\frac{1}{c}\) for the c in the big O definition)

\section{Examples with proofs}
% EXAMPLES WITH PROOFS

\paragraph{Claim 22.} \(n = O(2^n)\)
(This must seem very silly, but even though we have a strong feeling it’s true, we still need to learn how to PROVE it)
Proof. We will prove by induction that \( \forall n \ge 1 \) , \(2^n \ge n\), and that will suffice.
Basis: \(n = 1\), so it is clear that: \(n = 1 < 2 = 2^n\)
Assumption: Assume that \(n < 2^n\) for some \(n\).
Step: We will prove for \(n + 1\). It holds that:
\begin{equation*}
n + 1 < 2^{n} + 1 < 2^n + 2^n = 2^{n+1}
\end{equation*}
Now let’s try to analyze something we’re going to encounter in this course: Recursively defined functions (Like Fibonacci’s
sequence).
\paragraph{Example 23.} Let \(f(n)\) be defined as:
\begin{equation*}
f(n) = \left\{ \begin{array}{rcl}
& 2f\left( \lfloor  \frac{n}{2} \rfloor \right) + 1 & \mbox{for }  n > 1  \\
& 1 & \mbox{else}  
\end{array}\right.
\end{equation*}
Let’s find an asymptotic upper bound for \(f(n)\).
We can unravel this recursion rule, and to make things easier let’s just assume that n is a power of 2. So we get:
    \begin{equation*}
    f(n) = 2f\left(\frac{n}{2}\right) + 1 =2\left(2f\left(\frac{n}{4}\right) + 1\right) + 1 = ... = nf(1)+ \left( n -1 \right) = 2n-1
    \end{equation*}

Proof. \textbf{(Bad proof)}
Proof by strong induction. We’ll try to prove using \(N = 1, c = 10\), and we’ll try to show that \(f(n) < c \cdot n\).
Base: \(n = 1\). Clearly we get that \(f(1) = 1 < 10 = c \cdot 1\)

Assumption: Assume that for every \(m < n\), this claim holds.
Step: We get:
\begin{equation*}
    f(n) = 2f\left( \lfloor  \frac{n}{2} \rfloor \right) + 1 \le 2c\lfloor  \frac{n}{2} \rfloor  + 1 \le c\cdot n + 1
\end{equation*}
But this is not a proof... (why? Didn’t prove hypothesis)


Proof. \textbf{(Good proof)} We’ll try to prove something a little different. We’ll prove by strong induction that (with the same \(N, c\) ): \(f(n) < c \cdot n - 1\) And that will be enough (why? This implies \(f(n) = O(n)\)).
Base: \(n = 1\). Clearly, \(f(1) = 1 < 9 = c \cdot 1 - 1\)
Assumption: Assume that for every m < n, this claim holds.
Step: Then we get:
\begin{equation*}
    f(n) = 2f\left( \lfloor  \frac{n}{2} \rfloor \right) + 1 \le 2\left(c  \frac{n}{2} -1 \right)  + 1 \le c\cdot n - 1
\end{equation*}

Remark 24. Note that in the definition of big \(O\) notation (12), the property only needs to hold from some \(N\) and on. In this
last proof, we chose that \(N\) to be \(N = 1\), but it does not has to be 1.



\input{tex/texlib/tail}